---
title: "frab: An R Package for Adding Tables"
date: "2024-07-18"
abstract: >

  In R, adding two objects of class `table` has a natural
  interpretation.  However, in base R, adding two tables can give
  plausible but incorrect results.  The `frab` package provides a
  consistent and efficient way to add `table` objects, subject to
  `disordR` discipline [@hankin2022_disordR].  In this short article I
  introduce the `frab` package which allows one to add tables; it
  also  provides an alternative interpretation of named vectors.  The
  underlying mathematical structure is the Free Abelian
  group, hence "`frab`".
draft: true
author:  
  - name: Robin K. S. Hankin
    affiliation: University of Stirling
    address:
    - Compute and Mathematical Sciences, University of Stirling, Scotland
    url: https://www.stir.ac.uk/people/1966824
    orcid: 0000-0001-5982-0415
    email: hankin.robin@gmail.com
type: package
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: frab.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library("frab")
set.seed(0)
```


# Prologue: `table()`

Suppose we have three R tables:

```{r,label=badtables}
x <- table(c("a","a","b","c","d","d","a"))
y <- table(c("a","a","b","d","d","d","e"))
z <- table(c("a","a","b","d","d","e","f"))
```

Can we ascribe any meaning to `x+y` without referring to the arguments
sent to `table()`?  Well yes, we should simply sum the counts of the
various letters.  However:

```{r label=addtableswrong}
x
y
x+y
```

The sum is defined in this case.  However, close inspection shows that
the result is clearly incorrect.  Although entries for `a` and `b` are
correct, the third and fourth entries are not as expected: in this
case R idiom simply adds the entries elementwise with no regard to
labels.  We would expect `x+y` to respect the fact that we have 5 `d`
entries, even though element `d` is the fourth entry of `x` and the
third of `y`.  Further:

```{r, label=addtablesgiveserror,error=TRUE}
x
z
x+z
```

Above we see that `x` and `z` do not have a well-defined sum, in the
sense that `x+z` returns, quite reasonably, an error.

# Named vectors

A _named vector_ is a vector with a `names` attribute.  Each element
of a named vector is associated with a name or label.  The names are
not necessarily unique.  It allows you to assign a name to each
element, making it easier to refer to specific values within the
vector using their respective names.  Named vectors are a convenient
and useful feature of the R programming language [@rcore2022].
However, consider the following two named vectors:

```{r label=twonamedvectors}
x <- c(a=1,b=2,c=3)
y <- c(c=4,b=1,a=1)
```

Given that `x+y` returns a named vector, there are at least two
plausible values that it might give, viz:

`c(a=5,b=3,c=4)`

or

`c(a=2,b=3,c=7)`.

In the first case the elements of `x` and `y` are added pairwise, and
the `names` attribute is taken from the _first_ of the addends.  In
the second, the names are considered to be primary and the value of
each name in the sum is the sum of the values of that name of the
addends.  Note further that there is no good reason why the first
answer could not be `c(c=5,b=3,a=4)`, obtained by using the names
attribute of `y` instead of `x`.  The `frab` package provides a
consistent interpretation for named vectors in which addition is
indexed by name, not position.


## Package internals

The package uses the `STL map` class for efficiency.  This class
maps strings (symbols) to doubles; the declaration

```
typedef std::map <std::string, double> frab; 
```

appears in the `src/` package directory.  Such maps are limited
only by memory availability.

## The `frab` package in use

The `frab` package associates a numerical value with each of a set
of arbitrary (character string) symbols.  This is accomplished using
the `STL map` class, a container that stores key-value pairs
and allows fast lookup and insertion based on the key.  Here we have
keys as character strings and values are double-precision numbers.


The package uses a single `S4` class, `frab`, for which a variety of
methods is defined.  There are several ways to create `frab`
objects, but the most straightforward is to coerce a named vector
using the `frab()` function:

```{r frabzyx}
frab(c(z=2,y=7,x=1))
```

Above, see how `frab()` takes a named numeric vector and returns
an object of class `frab`.  It takes the names of its argument,
possibly reordering them, and returns a `frab` object.  Function
`frab()` considers the names of the elements to be the primary
extraction and replacement mechanism.  If the argument has repeated
names, function `frab()` sums them:

```{r frab_show_addition}
frab(c(t=3,q=2,t=4,q=-1,p=6,a=3,t=5))
```

Above we see that the entries for `t` and `q` are summed.
Zero entries are discarded:

```{r frab_show_zero_discarded}
frab(c(pear=1,kiwi=0,fig=3,lime=2,fig=-3))
```


Above we see that zero entries [`kiwi` and `fig`] are discarded,
irrespective of whether a zero is explicitly given, or repeated values
cancel.  However, the main motivation for using `frab` objects is
that they may be added:

```{r howfrabaddition}
a <- frab(c(x=2,y=1,z=3))
b <- frab(c(y=3,x=3,u=1))
a+b
```

Above we see that the sum of `a` and `b` is defined in terms of the
keys of the summands, not position. Thus, `a+b` has 5 ($=2+3$) for its
`x` entry and 4 ($=1+3$) for its `y` entry.  Observe that `a+b` has
its entries in implementation-specific order, as per `disordR`
discipline.  Further, note that `a+b` has length 4, while `a` and `b`
have length 3


# Extensions of `frab` objects to floating-point values

The `frab` class is sufficiently flexible to incorporate
floating-point values, although one has to be a little careful with
numerical round-off errors:

```{r showfrabroundofferrors}
x <- frab(c(a=4,u=pi,p=exp(pi)))
y <- frab(c(p=-exp(pi)/3,u=-pi))
z <- frab(c(p=-exp(pi)*2/3))
x+y+z
```

Above, the sum `x+y+z` should have zero entries for `u` and `p`; we
see the entry for `u` vanishes, but the entry
for `p` is nonzero, being subject to (small) numerical roundoff
error.

# The `frab` package and `disordR` discipline

The `frab` package conforms to disord discipline
[@hankin2022_disordR].  Here I present some discussion of the
motivation for this design decision.  Consider the following short R
session:

```{r shortsession, error = TRUE}
(a <- frab(c(x=2,y=1,u=8,z=3,v=5)))
a["x"]
a[1]
```

Above we see a `disordR` violation: there is no such thing as
"`a[1]`", for the elements of `a` are stored in an
implementation-specific order.  The `STL map` class stores value-key
pairs in an undefined order; thus, extracting the value for `x` is
fine, but because the order is not defined it makes no sense to
extract the ``first" element (and attempting to do so results in a
disord discipline error).

Observe that we cannot dispense with order of the values entirely,
because sometimes I am interested in the vector of keys, or their
values, in isolation.  If we want to work with the names or values of
a `frab` object, then the `disord` print methods are used:

```{r frabnamesisolation}
a <- frab(c(x=2,y=1,z=3))
names(a)
values(a)
```

Above we see that `names(a)` and `values(a)` return `disord` objects,
in this case with the same hash code which indicates that the objects
are consistent with one another in the sense of
`disordR::consistent()`.  These objects may be displayed and
subsequently manipulated, subject to disord discipline:

```{r frabnamereplace}
(a <- frab(c(x=2,y=1,z=3)))
names(a) <- toupper(names(a))
a
```

We see that replacement methods are defined, in this case `toupper()`
returns a `disord` object and the names of `a` become their uppercase
equivalents.

Again observe that there is no meaning to the operation ``extract the
first element of `names(a)`", because the elements of `names(a)`,
being a `disord` object, are stored in an implementation-specific
order.  We may manipulate the values of a `frab` object, if we are
careful to be consistent with disord discipline.  The package includes
a number of convenient replacement idioms:

```{r frabvaluemanipulate}
(a <- frab(c(x=2,y=-1,z=3,p=-4,u=20)))
values(a) <- values(a)^2
a
```

Thus, we may square the `values()` of `a` using the replacement
method, and object `a` is altered appropriately.  Further, we may use
the `disindex` class of the `disordR` package to replace certain
values using standard square bracket replacement idiom:

```{r frabvaluemanipulate2}
(a <- frab(c(x=2,y=-1,z=11,p=-4,u=20)))
a[a>10] <- 19
a
```

Thus, we set any value exceeding 10 to 19.

# Higher-dimensional tables

The ideas above have a natural generalization to multi-dimensional
tables.  The relevant `frab` package `S4` class is `sparsetable`, and
this has arithmetic methods implemented.  We start with
three-dimensional tables (two-dimensional `sparstable` objects have a
print method that coerces them to arrays before displaying, and the
general case is easier to understand).

```{r threedeeexample}
abcd <- letters[1:4]
jj <- sample(abcd,99,repl=T,prob=1/(1:4)^2)
I <- matrix(jj,33,3)
head(I)
nrow(I)
(x <- sparsetable(I))
```


Thus, random matrix `I` has has 33 rows and elements a-d .Coercing `I`
to `sparsetable` object `x` shows the counts of each type of row.
Object `x` has 13 rows; not all $4^3=64$ possible row types are
represented.  Standard extraction and replacement operations work as
expected on `sparsetable` object `x`:

```{r standardextraction}
x["a","a","b"]
x[rbind(c("a","b","b"),c("a","b","c"))]
```

Thus, extraction can operate with the arguments sent separately, or as
a matrix.  However, other more specialised extraction methods are also
provided:

```{r disordextraction}
x[x>3]
```


Thus, extraction can operate with a disord object; here we extract
every element greater than 3.  Replacement methods are also defined:

```{r disordreplacement}
x[x==1] <- 0
x
```

Also, replacement methods can add new entries if needed:

```{r disordreplacementnew}
x["a","x","y"] <- 1000
x
```

Above we create a new entry corresponding to `a x y` with value
$1000$; the new entry is placed in its own implementation-specific
location, as per `disordR` discipline.


# Two dimensional tables

If a `sparsetable` object has arity 2, it is coerced to matrix
form before printing.  Otherwise, the semantics are the same as for
any other `sparsetable` object.

```{r twodeetable}
(x <- rspar2(9))
(y <- rspar2(9))
x+y
```

Above, `rspar()` gives a random 2D sparse table.  Package idiom allows
sparse tables to be added even if the tables are different sizes; the
resulting sum is automatically resized to accommodate both
addends. Entries with nonzero values in both `x` and `y` are
correctly summed.




# Appendix: the Free Abelian Group

The _Free Abelian group_ is an interesting and useful mathematical
object [@cohn1984]: it is a direct sum of infinite cyclic groups.  If
these cyclic goups are generated by $\left\lbrace x_i\colon
i\in\mathcal{I}\right\rbrace$ for some (finite) index set
$\mathcal{I}$, then the Free Abelian group $F$ will be

\begin{equation}\label{fab_def}
F=\bigoplus_{i\in\mathcal{I}}\left\langle x_i\right\rangle.
\end{equation}

Because $\left|\mathcal{I}\right| =k < \infty$, elements of $F$ will
be of the form

\begin{equation}\label{formalform}
g=n_1x_1+n_2x_2+\cdots+n_kx_k
\end{equation}

where $n_i\in\mathbb{Z}$, $1\leqslant i\leqslant k$.  The group
operation (conventionally one uses additive notation) is then defined
by componentwise addition.  If

\begin{equation}\nonumber
g=n_1x_1+n_2x_2+\cdots+n_kx_k\qquad
h=r_1x_1+r_2x_2+\cdots+r_kx_k,
\end{equation}

then 
\begin{equation}\label{formalhplusg}
h+g=(n_1+r_1)x_1+(n_2+r_2)x_2+\cdots+(n_k+r_k)x_k
\end{equation}

One can define $F$ formally by starting with a generating set
$X=\left\lbrace x_1,\ldots,x_k\right\rbrace$ of symbols.  Then $F$ is
the set of all formal expressions of the form \ref{formalform} under
the addition defined in equation \ref{formalhplusg}; the group axioms
follow immediately.  Alternatively, one can adopt the somewhat more
abstract category-theoretic approach.  For any set $S$ the _free group
on $S$_ is a group such that for any Abelian group $G$ and map
$f\colon S\longrightarrow G$, there exists a unique homomorphism
$\phi$ such that the following diagram commutes:

```{r echo=FALSE}
knitr::include_graphics("./commuting_diagram.pdf")
```

For general (non-Abelian) free groups in the context of R, see the
`freegroup` package [@hankin2022_freegroup_rnews].

# References
